---
title: "Résultats DID"
format: 
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    fig-pos: 'h'
execute: 
  echo: FALSE
---


```{r, include=FALSE}
library(tidyverse)
library(patchwork)
library(gt)



```

# Tableaux récapitulatifs

## PAC air/eau
```{r}
tribble(~Revenus,~CEE,~Prix, ~CroissanceNgestes,
        "Très modestes", "-", "+","+",
        "Modestes", "+*", "-","+*",
        "Intermédiaires", "+*","-","+*",
        "Remarques", "Pour TMO : effet signif en % prix ttc","",""
        ) |> 
  gt()
```

## Chaudière à granulés

```{r}
tribble(~Revenus,~CEE,~Prix, ~CroissanceNgestes,
        "Très modestes", "+*", "+","+",
        "Modestes", "-*", "-","+",
        "Intermédiaires", "-","+","+",
        "Remarques", "","",""
        ) |> 
  gt()
```

# Graphiques

## PAC

### Distributions pre/post

![](graphs_19_06_2025/cee/tmo/graph_ze_cee_tmo_density_Chaudière à granulés.pdf)

![](graphs_19_06_2025/cee/mo/graph_ze_cee_mo_density_Chaudière à granulés.pdf)

![](graphs_19_06_2025/cee/int/graph_ze_cee_int_density_Chaudière à granulés.pdf)

### CEE

**ZE/mois**
![](graphs_19_06_2025/results/pcae/graph_did_monthly_pcae_tmo_cee.pdf)

**Commune/mois**

![](graphs_19_06_2025/results/pcae/graph_did_monthlycom_pcae_tmo_cee.pdf)

**ZE/trimestre**

![](graphs_19_06_2025/results/pcae/graph_did_qtr_pcae_tmo_cee.pdf)

**Estimateur DCDH**

![](graphs_19_06_2025/results/pcae/graph_did_monthly_dcdhcsa_pcae_tmo_cee.pdf)

### Prix

**ZE/mois**
![](graphs_19_06_2025/results/pcae/graph_did_monthly_pcae_tmo_relprice.pdf)

**Commune/mois**

![](graphs_19_06_2025/results/pcae/graph_did_monthlycom_pcae_tmo_relprice.pdf)

**ZE/trimestre**

![](graphs_19_06_2025/results/pcae/graph_did_qtr_pcae_tmo_relprice.pdf)

**Estimateur DCDH**

![](graphs_19_06_2025/results/pcae/graph_did_monthly_dcdhcsa_pcae_tmo_relprice.pdf)

## Chaudière à granulés

**ZE/mois**
![](graphs_19_06_2025/results/cag/graph_did_monthly_cag_tmo_cee.pdf)

**Commune/mois**

![](graphs_19_06_2025/results/cag/graph_did_monthlycom_cag_tmo_cee.pdf)

**ZE/trimestre**

![](graphs_19_06_2025/results/cag/graph_did_qtr_cag_tmo_cee.pdf)

**Estimateur DCDH**

![](graphs_19_06_2025/results/cag/graph_did_monthly_dcdhcsa_cag_tmo_cee.pdf)

### Prix

**ZE/mois**
![](graphs_19_06_2025/results/cag/graph_did_monthly_cag_tmo_relprice.pdf)

**Commune/mois**

![](graphs_19_06_2025/results/cag/graph_did_monthlycom_cag_tmo_relprice.pdf)

**ZE/trimestre**

![](graphs_19_06_2025/results/cag/graph_did_qtr_cag_tmo_relprice.pdf)



